Joint photometric modeling for improved measurements in imaging surveys.

### 2014-12-02 documents/paper/ms_v0.0.tex
some intro; pseudocode

\documentclass[12pt,preprint]{aastex}
\usepackage{amssymb,amsmath,mathrsfs,hyperref,datetime}

\newcommand{\datavector}[1]{\boldsymbol{#1}}
\newcommand{\data}{\datavector{y}}
\newcommand{\datum}{\data_i}
\newcommand{\truedatum}{\data_{{\rm true}, i}}
\newcommand{\noiselessdata}{\tilde{\data}}
\newcommand{\noiselessdatum}{\tilde{\datum}}
\newcommand{\xdcov}{\datavector{V}_j}
\newcommand{\datacov}{\datavector{C}}
\newcommand{\postcov}{\tilde{\datacov}_{i, j}}
\newcommand{\vmu}{\datavector{\mu}}
\newcommand{\postmu}{\tilde{\vmu}_{i, j}}
\newcommand{\datumcov}{\datacov_i}
\newcommand{\transpose}{{\rm T}}
\newcommand\ttt[1]{{\texttt{#1}}}
\newcommand\rf[1]{{\bf [RF: #1]}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Section - Method
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Extreme Deconvolution}

@@

@@

\begin{eqnarray}\displaystyle
p(\noiselessdata) = \sum_j^K \frac{\alpha_j}{\sqrt{2\pi\det{|\xdcov|}}} \exp \left (-\frac{1}{2}(\noiselessdata - \vmu_j)^\transpose \xdcov^{-1}(\noiselessdata - \vmu_j) \right )
\quad .
\label{eqn:noiseless}
\end{eqnarray}

@@

\begin{eqnarray}\displaystyle
p(\truedatum | \datum) & \propto & \sum_j^K \frac{\alpha_j}{\sqrt{2\pi\det{|\postcov|}}} \exp \left (-\frac{1}{2}(\truedatum - \postmu)^\transpose \postcov^{-1}(\truedatum - \postmu) \right ) \\
{\rm where,}\\
\postcov & = & (\datumcov^{-1} + \xdcov^{-1})^{-1} \quad {\rm and,} \\
\postmu & = &  \postcov^{-1} (\datumcov^{-1} \datum + \xdcov^{-1} \vmu_j)
\label{eqn:posterior}
\end{eqnarray}

\section{Data and Model Considerations}

@@

\section{Results and Discussion}

@@

\begin{figure}
\centering
\includegraphics[clip=true, trim=1.5cm 0.5cm 1.5cm 0.5cm,
  width=16cm]{fig2.png}
\caption{Error ellipses for single epoch S82 photometry (dashed) and XD
posterior distributions for a star (top row) and galaxy (bottom row).  The 
projections are the same as in Figure \ref{fig:contours}.  Ellipses for the XD
posteriors are generally smaller, yet are still consistent with the original 
measurements.
}
\label{fig:posteriors}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[clip=true, trim=1.5cm 0.5cm 1.5cm 0.5cm,
  width=16cm]{fig3.png}
\caption{Plotted are median values for the 1 sigma uncertainty in $r$ PSF
magnitude, $r$
psf minus model magnitude, and the four model colors used.  The solid black 
curve is for the Stripe 82 coadd, while the dotted black and red curves are for
single epoch Stripe 82 and DR10 data, respectively.  The dashed curves show the
median uncertainty for the XD posteriors.  While not quite as precise as the 
coadd data, the XD posteriors are a significant improvement over the catalog
data.
}
\label{fig:error_rates}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[clip=true, trim=1.5cm 0.5cm 1.5cm 0.5cm,
  width=16cm]{fig4.png}
\caption{In the top row, we plot 2D histograms of $r$ PSF magnitudes, $u-g$
model color, and $i-z$ model color for single epoch Stripe 82 data versus the
corresponding coadd values.  In the bottow row, we plot the same for a single
sample from the XD posteriors of the data in the top.  Generally speaking the
XD posteriors are very consistent with the coadd photometry, indicating that 
the models are indeed acurate.
}
\label{fig:xx_plots}
\end{figure}

>>>>

\newcommand{\xdmean}{\datavector{\mu}}
\documentclass[12pt,preprint]{aastex}
\usepackage{amssymb,amsmath,mathrsfs,hyperref,datetime}

\usepackage{algorithmic,algorithm}
\usepackage{eqparbox}
\renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENT}
                                      {\it~#1}}
\newcommand{\blankcomment}{\protect\phantom{$1$}\hspace{10cm}
                           \protect\phantom{$1$}}

\newcommand{\datavector}[1]{\boldsymbol{#1}}
\newcommand{\data}{\datavector{y}}
\newcommand{\datum}{\data_i}
\newcommand{\truedatum}{\data_{{\rm true}, i}}
\newcommand{\noiselessdata}{\tilde{\data}}
\newcommand{\noiselessdatum}{\tilde{\datum}}
\newcommand{\xij}{\datavector{x}_{ij}}
\newcommand{\Lij}{\datavector{\lambda}_{ij}}
\newcommand{\Tij}{\datavector{T}_{ij}}
\newcommand{\qij}{\datavector{q}_{ij}}
\newcommand{\qj}{\datavector{q}_{j}}
\newcommand{\dlt}{(\xdmu_j - \xij)}
\newcommand{\xdcov}{\datavector{V}_j}
\newcommand{\datacov}{\datavector{C}}
\newcommand{\postcov}{\tilde{\datacov}_{i, j}}
\newcommand{\xdmu}{\datavector{\mu}}
\newcommand{\fixedmu}{\bar{\mu}}
\newcommand{\postmu}{\tilde{\xdmu}_{i, j}}
\newcommand{\datumcov}{\datacov_i}
\newcommand{\transpose}{\mathsf{T}}
\newcommand\ttt[1]{{\texttt{#1}}}
\newcommand\rf[1]{{\bf [RF: #1]}}

@@

For a large catagory of survey missions, the main scientific deliverable is a
catalog of photometric measurements such as astrometry, magnitudes, and
morphological information.  These catalogs are almost always constructed on a 
per-object basis -- each source in an image is measured 
independently\footnote{Apart from issues associated with blending, detection,
psf modeling, etc.}.
For most types of sources (stars, galaxys and quasars), we have thousands to
millions of measurements albeit with varying properties and signal-to-noise.
This begs the question as to whether we can make a better estimate of a given
photometric measurement based on the fact that we have many similar
measurements.  Indeed, if we have thousands of measurements of (for example)
main sequence F4 stars, we ought to be able to get a better estimate of a new 
ones photometric properties by incorperating what is learned from the
population.

Since telescope time is expensive both in terms of cost and person-hours, it
behoves us to consider whether we can make better estimates of photometric 
measurements at (essentially) fixed expense.  We argue that it is indeed
possible to improve the measurements, by exploiting the similarity across
observations.  As a demonstration, we use Extreme Deconvolution
\citep[XD, ][]{bovy09} to model the photometry in the SDSS DR10 \citep{} and
Stripe 82 \citep{} data releases.  XD has been shown to be a powerful yet
simple way to model the properties of heteroskedastic data \citep{bovy10}.
Here, we apply it to SDSS photometry and produce posterior estimates of the 
measurements.  In Section \ref{sec:method} we discuss the ideas and methods 
behind our modeling, Section \ref{sec:data} describes the data we use for 
our demostration, and \ref{sec:results} presents and discusses the application
of our XD model.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Section - Method
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Posterior Estimates with Extreme Deconvolution}
\label{sec:method}

@@

\begin{algorithm}
\caption{Expectation-Maximization for XD}
\textbf{Expectation Step:}
\begin{algorithmic}[1]
\STATE $\Tij \gets \xdcov + \datumcov$
\STATE $\qij \gets \frac{\alpha_j \mathcal{N}(\datum | \xdmu_j, \Tij)}
       {\sum_j \alpha_j \mathcal{N}(\datum | \xdmu_j, \Tij)}$
\STATE $\xij \gets \xdmu_j + \xdcov\Tij^{-1}(\datum - \xdmu_j)$
\STATE $\Lij \gets \xdcov - \xdcov\Tij^{-1}\xdcov$
\end{algorithmic}
\textbf{Maximization Step:}
\begin{algorithmic}[1]
\STATE $\alpha_j \gets \frac{1}{N}\sum_i \qij$
\STATE $\xdmu_j \gets \frac{1}{\qj}\sum_i \qij\xij$
\FOR [\blankcomment]{$k = 1, \ldots, N_{\rm dim}$}
    \IF [// fix mean of feature k to $\fixedmu_{jk}$ if 
        desired]{$\exists\, \fixedmu_{jk}$}
        \STATE $(\xdmu_j)_k \gets \fixedmu_{jk}$
    \ENDIF
\ENDFOR
\STATE $\xdcov \gets \frac{1}{\qj}\sum_i \qij [\dlt\dlt^\transpose + \Lij]$
\FOR {$k = 1, \ldots, N_{\rm dim}$}
    \IF [// axis align covariance for feature k if desired]
        {$align_{jk}$}
        \STATE $var = (\xdcov)_{k,k}$
        \STATE $(\xdcov)_{k,:} \gets 0$
        \STATE $(\xdcov)_{:,k} \gets 0$
        \STATE $(\xdcov)_{k,k} \gets var$
    \ENDIF
\ENDFOR
\end{algorithmic}
\end{algorithm}

@@

\begin{eqnarray}\displaystyle
p(\noiselessdata) = \sum_j^K \frac{\alpha_j}{\sqrt{2\pi\det{|\xdcov|}}} \exp \left (-\frac{1}{2}(\noiselessdata - \xdmu_j)^\transpose \xdcov^{-1}(\noiselessdata - \xdmu_j) \right )
\quad .
\label{eqn:noiseless}
\end{eqnarray}

@@

\begin{eqnarray}\displaystyle
p(\truedatum | \datum) & \propto & \sum_j^K \frac{\alpha_j}{\sqrt{2\pi\det{|\postcov|}}} \exp \left (-\frac{1}{2}(\truedatum - \postmu)^\transpose \postcov^{-1}(\truedatum - \postmu) \right ) \\
{\rm where,}\\
\postcov & = & (\datumcov^{-1} + \xdcov^{-1})^{-1} \quad {\rm and,} \\
\postmu & = &  \postcov^{-1} (\datumcov^{-1} \datum + \xdcov^{-1} \xdmu_j)
\label{eqn:posterior}
\end{eqnarray}

\section{Data and Model Considerations}
\label{sec:data}

@@

\section{Results and Discussion}
\label{sec:results}

@@

\begin{figure}
\centering
\includegraphics[width=16cm]{fig2.png}
\caption{Error ellipses for single epoch S82 photometry (dashed) and XD
posterior distributions for a star (top row) and galaxy (bottom row).  The 
projections are the same as in Figure \ref{fig:contours}.  Ellipses for the XD
posteriors are generally smaller, yet are still consistent with the original 
measurements.
}
\label{fig:posteriors}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=16cm]{fig3.png}
\caption{Plotted are median values for the 1 sigma uncertainty in $r$ PSF
magnitude, $r$
psf minus model magnitude, and the four model colors used.  The solid black 
curve is for the Stripe 82 coadd, while the dotted black and red curves are for
single epoch Stripe 82 and DR10 data, respectively.  The dashed curves show the
median uncertainty for the XD posteriors.  While not quite as precise as the 
coadd data, the XD posteriors are a significant improvement over the catalog
data.
}
\label{fig:error_rates}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=16cm]{fig4.png}
\caption{In the top row, we plot 2D histograms of $r$ PSF magnitudes, $u-g$
model color, and $i-z$ model color for single epoch Stripe 82 data versus the
corresponding coadd values.  In the bottow row, we plot the same for a single
sample from the XD posteriors of the data in the top.  Generally speaking the
XD posteriors are very consistent with the coadd photometry, indicating that 
the models are indeed acurate.
}
\label{fig:xx_plots}
\end{figure}
